Elaboration with First-Class Implicit Function Types

### 2020-02-20 paper.tex
stuff

\begin{definition}[Closing type] For each $\Theta|\Gamma \vdash A : \U$, we define
$\Gamma \To A$ by recursion on $\Gamma$, such that $\Theta|\emptycon\vdash
  \Gamma \To A : \U$.
  \begin{alignat*}{3}
    &((\Gamma,\, x : A) \To B)     && :\equiv (\Gamma \To ((x : A) \to B))\\
    &((\Gamma,\, x : A = t) \To B) && :\equiv (\Gamma \To B[x\mapsto t])\\
    &(\emptycon \To B)             && :\equiv B
  \end{alignat*}
\end{definition}

@@

  &\quad\eif\,((x : A) \in \Gamma)\,\lor\,((x : A = t) \in \Gamma) \\

@@

\begin{theorem}[Conservativity]
Elaboration is conservative over the surface syntax, in the sense that for any
surface term $t$, if checking or inference outputs $t'$, then the forgetful
translation of $t'$ differs from $t$ only by
  \begin{itemize}
    \item Having all $\_$ holes filled in by expressions.
    \item Having extra implicit $\lambda$-s and implicit applications inserted.
  \end{itemize}
This follows by straightforward induction on surface syntax. \qed
\end{theorem}

@@

\begin{itemize}
  \item \emph{Let-generalization}. This is an open
    research topic in settings with dependent types, and we make no attempt at
    tackling it. See \cite{eisenberg2016dependent} for a treatment in a proposed
    dependent version of Haskell.
  \item \emph{Polymorphic subtyping}. In some prior works, e.g.\ in
    \cite{dunfield2013complete, vytiniotis2008fph}, there is a subtyping
    relation arising along instantiations of polymorphic types. In GHC 8
    polymorphic subtyping is implemented for function types only. Polymorphic
    subtyping complicates type inference, and to our knowledge it has not been
    implemented in any dependently typed setting. We also believe that it is an
    anti-feature in dependent settings, because the elaboration of subtyping
    must insert coercions which significantly change the intensional character
    of programs. For example, if we have covariant list types, then coercing $t
    : List\,(\{A : \U\}\to A \to A)$ to $t : List\,(Bool \to Bool)$ requires
    mapping over $t$ and inserting implicit applications to $Bool$ for each list
    element. In System F, this is harmless, since types are computationally
    irrelevant, but in our core syntax we have implicit functions with arbitrary
    (relevant) domains. In GHC, subtyping coercions for functions change
    operational semantics; this is a reason for abandoning subtyping in recent
    developments of impredicative inference for GHC \cite{serrano2020a}.
\end{itemize}

\section{Issues with First-Class Implicit Functions}

@@

%% Π : (Δ : Tel) → (Rec Δ → Set) → Set
%% Π ε       B = B tt
%% Π (A ▶ Δ) B = (x : A) → Π (_ , Δ x) λ δ → B (x , δ)

%% app : ∀ {Δ B} → Π Δ B → (δ : Rec Δ) → B δ
%% app {ε}     f x         = f
%% app {A ▶ Δ} f (x₀ , x₁) = app (f x₀) x₁

%% lam : ∀ {Δ B} → ((δ : Rec Δ) → B δ) → Π Δ B
%% lam {ε}     f = f tt
%% lam {A ▶ Δ} f = λ x → lam λ δ → f (x , δ)

>>>>

\begin{definition}[Closing type]\label{def:closingtype} For each $\Theta|\Gamma \vdash A : \U$, we define
$\Gamma \To A$ by recursion on $\Gamma$, such that $\Theta|\emptycon\vdash
  \Gamma \To A : \U$.
  \begin{alignat*}{3}
    &((\Gamma,\, x : A) \To B)     && :\equiv (\Gamma \To ((x : A) \to B))\\
    &((\Gamma,\, x : A = t) \To B) && :\equiv (\Gamma \To B[x\mapsto t])\\
    &(\emptycon \To B)             && :\equiv B
  \end{alignat*}
\end{definition}

@@

  &\quad\eif\,(\Gamma = (\Gamma_0,\,x : A,\,\Gamma_1))\,\lor\,(\Gamma = (\Gamma_0,\,x : A = t,\,\Gamma_1)) \\

@@

\begin{theorem}[Conservativity]\label{thm:conservativity}
Elaboration is conservative over the surface syntax, in the sense that for any
surface term $t$, if checking or inference outputs $t'$, then the forgetful
translation of $t'$ differs from $t$ only by
  \begin{itemize}
    \item Having all $\_$ holes filled in by expressions.
    \item Having extra implicit $\lambda$-s and implicit applications inserted.
  \end{itemize}
This follows by straightforward induction on surface syntax. \qed
\end{theorem}

@@

\begin{itemize}
  \item \emph{Let-generalization}. This is an open research topic in settings
    with dependent types, and we make no attempt at covering it. See
    \cite{eisenberg2016dependent} for a treatment in a proposed dependent
    version of Haskell.
  \item \emph{Polymorphic subtyping}. In some prior works, e.g.\ in
    \cite{dunfield2013complete, vytiniotis2008fph}, there is a subtyping
    relation arising along instantiations of polymorphic types. In GHC 8
    polymorphic subtyping is implemented for function types only. Polymorphic
    subtyping complicates type inference, and to our knowledge it has not been
    implemented in any dependently typed setting. We also believe that it is an
    anti-feature in dependent settings, because elaboration of subtyping must
    insert coercions which significantly change the intensional character of
    programs. For example, if we have covariant list types, then coercing $t :
    List\,(\{A : \U\}\to A \to A)$ to $t : List\,(Bool \to Bool)$ requires
    mapping over $t$ and inserting implicit applications to $Bool$ for each list
    element. In System F, all such coercions are erasible, since types are
    computationally irrelevant, but in our core syntax we have implicit
    functions with arbitrary (relevant) domains. In GHC, subtyping coercions for
    functions change operational semantics; this is a reason for abandoning
    subtyping in recent developments of impredicative inference for GHC
    \cite{serrano2020a}.
\end{itemize}

\section{Issues with First-Class Implicit Functions}
\label{sec:issues}

@@

\section{Extending Elaboration}

We shall utilize the extended core theory to implement smarter
elaboration. Recall from Section \ref{sec:issues} that the old elaborator makes
two kinds of unforced insertion choices:
\begin{enumerate}
\item $\echeck{t}{\Theta}{\Gamma}{A}$ does not insert an implicit $\lambda$ when $A$ is meta-headed.
\item $\einsert{\true}{A}$ does not insert an implicit application if $A$ is meta-headed.
\end{enumerate}
In the following, we shall only handle the case for $\lambda$ insertions. This
allows a simple implementation which only requires minimal changes to
unification, and which is already remarkably powerful. It seems that handling
implicit application insertions requires extending unification; we will talk
about this in Section \ref{TODO}. But before delving into elaboration, we need to
rectify a curious issue.

\subsection{Superfluous Implicit Functions}

Assuming $Bool : \U$, $true : Bool$ and $false : Bool$, consider the following surface expression:
\[
  \slet\,x : \_ = true\,\sin\,x
\]
What should this expression elaborate to? We would expect the result to be simply
\[
  \slet\,x : Bool = true\,\sin\,x
\]
However, there are infinitely many core terms which are conservative over the
surface expression in the sense of Theorem \ref{thm:conservativity}. That is, we
can wrap definitions with any number of implicit $\lambda$-s, and add implicit
applications accordingly to usage sites of the defined name. For example, we could have
\[
  \slet\,x : \{y : Bool\} \to Bool = \lambda\,\{y\}.\,true\,\sin\,x\,\{true\}
\]
This is clearly undesirable. With the type $\{y : Bool\} \to Bool$, the implicit
argument $y$ is never inferable, because the codomain type does not depend on
the domain, and the argument is never constrained. Hence, with the above
definition, we always have to write $x\,\{true\}$ or $x\,\{false\}$ when we want
to use $x$. To rule out such nonsense, we adopt the following principle:

\begin{quote}
  \emph{Elaboration should never invent non-dependent implicit function types.}
\end{quote}

This was a non-issue in the old elaborator, because it was not able to invent
implicit function types; it was only utilizing the type annotations present in
the surface input. With $\slet\,x : \_ = true$, the old elaborator checks $true$
with a fresh meta, and just assumes that the meta does not stand for an
implicit function type.

To implement our principle, we need \emph{constancy constraints}. These are
constraints in the sense of GHC's OutsideIn(X) \cite{vytiniotis2011outsidein}
and also in the sense of the implementation in Agda and Coq. Rules for constancy
constraints are as follows.

\begin{mathpar}
  \inferrule*[lab=metacon/constr]
             {\Theta \vdash \alpha : \Gamma \To \Tel \\ \Theta|\Delta,\,x : \Rec\,(\alpha\,\ol{t}) \vdash B : \U}
             {\Theta,\,\mathsf{constancy}_{\Delta,\,x : \Rec\,(\alpha\,\ol{t})}\,B\vdash}

  \inferrule*[lab=constr/solve]
             {\Theta_0 \vdash \alpha : \Gamma_0 \To \Tel \\ x \notin FreeVars(B)}
             {\Theta_0,\,\mathsf{constancy}_{\Delta,\,x : \Rec\,(\alpha\,\ol{t})}\,B,\,\Theta_1|\Gamma_1
               \vdash \alpha \equiv \lambda\,\overline{\mathsf{vars}_{\Gamma_0}}.\,\epsilon}
\end{mathpar}

In the rule \LabTirName{metacon/constr}, we allow extension of the metacontext
with a constraint.  In the constraint, the metavariable $\alpha : \Gamma \To
\Tel$ represents an unknown domain for a curried function type, and $B$ is the
codomain of the function. Here, $\Gamma \To \Tel$ refers to a closing type, which
was defined in Section \ref{def:closingtype}.

\todo{first, define new closing types and fresh metas!}

The rule \LabTirName{constr/solve} says that if $B$ does not depend on the $x :
\Rec\,(\alpha\,\ol{t})$ domain variable, then $\alpha$ is equal to the constant
function which returns the empty telescope. We use the abbreviation
$\lambda\,\overline{\mathsf{vars}_{\Gamma_0}}$ to refer to an iterated $\lambda$
which abstracts over all variables in $\Gamma_0$.
